---
title: "ERA 5 calibration to Elexon power"
author:
  - "S. Gomez$^1$"
institute:
   - $^1$School of Mathematics, University of Edinburgh
institute-short: "UoE"
date: "2025-12-05"
format: 
  beamer:
    theme: Rochester
    colortheme: spruce
    fonttheme: professionalfonts
    fig-cap-location: top
    toc: true
    slide-level: 2
    split-bib: true
    output-file: "era5_calibration_slides.pdf"
header-includes:
  - \usepackage{graphicx}
fontsize: 12pt
aspectratio: 169
editor: source
nocite: "@*"
---

```{r include=FALSE}
require(tidyverse)
require(kableExtra)
require(parallel)
require(dplyr)
require(scales)
require(ggsci)

require(jsonlite)
require(httr)

require(data.table)
require(sf)
library(rnaturalearth)
library(rnaturalearthdata)

require(plotly)
require(ggthemes)
require(arrow)
path <- "~/Documents/elexon"
```

```{r}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo = FALSE, cache = FALSE)
```

# Calibration to power {#sec-intro}

## Updates {#sec-updates .unnumbered .nolink}

### New
- Spatial correlation check
- Updated linear models
- Power curve modelling

### Next steps
- Combine PC model with calibration
- Benchmark models: Quantile mapping, GAMs
- Calibration spatiotemporal model



# Overview {#sec-appendix}

## Data sources 
::: {.columns}

::: {.column width="50%"}
### Wind speed
- ERA5 at wind farms
  - Hourly data
  - Spatial resolution 0.25° x 0.25° 
  - 10m and 100m heights

:::

::: {.column width="50%"}
### Wind power
- Elexon BMU data (since 2019)
  - Half hourly data
  - Generation, curtailment, potential, capacity
  - Outage data (REMIT)
- REPD database
  - Location, turbine height, capacity
:::

:::

## Overview {#sec-overview}
### 1. ERA 5 to wind farm 
Vertical interpolation to turbine height $h$.

\centering
$w(h) = w_{100} \left( \frac{h}{100} \right)^{\alpha}$, 
where $\alpha = 1/7$

### 2. Wind speed to power
Generic power curves rescaled to wind farm capacity.
\begin{equation*}
\hat{PC}_i(w) = PC_i(w)\times \frac{C_i}{\text{Rated power}},
\end{equation*}

where $C_i$ is the capacity at location $i$


## Overview of power conversion
### 3. Potential generation
Curtailment and outages are two main events that impact observed generation $o_{it}$

- Curtailment is added giving rise to potential generation:
\begin{equation*}
p_{it} = o_{it} + \text{curt}_{it}
\end{equation*}

- Outage data shows additional limits on capacity
- Currently outage periods are excluded

## Calibration

### 4. Calibration
ERA5-derived power estimate $\hat p_{it}$ is compared versus potential power $p_{it}$
\begin{align*}
\hat p_{it} &= \hat{PC}_i(w_{it})\\
p_{it} &= \beta_0 + \beta \hat p_{it} + s_i + u_t,
\end{align*}

where $s_i$ and $u_t$ represent spatial and temporal effects.

# Calibration with linear model

## GB level aggregation
::: {.columns}

::: {.column width="50%"}

- Power aggregated at GB level
  \begin{align*}
  p_{\text{tot},t} &= \sum_i p_{it} / \sum_i C_i 
  \end{align*}
- Each point represents one hour
- Overestimation persits but dispersion is lower now
:::

::: {.column width="50%"}
\centering
\small Scatter aggregated power
\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/gb_aggr_scatter_all.png}
:::

:::

## Initial linear model

::: {.columns}

::: {.column width="50%"}

Features of importance: 
- month
- Turbine type (offshore / onshore)

\begin{equation*}
p^{calib}_t = \alpha_{k,m} + \beta_{k,m} \hat p_{t},
\end{equation*}
where $\alpha_{k,m}$ and $\beta_{k,m}$ represent the intercept and slope, varying by type and month

:::

::: {.column width="50%"}
\centering
\small Scatter aggregated power
\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/gb_calib_bindens_all.png}
:::

:::

# Previous research {#sec:literature}

## Power curve modelling

::: {.columns}

::: {.column width="50%"}

### Binning method
\begin{equation*}
  P_i = \frac{1}{n_i} \sum_{j=1}^{n_i} P_{ij}
\end{equation*}
where: $P_{ij}$ is the $j$ th power observation in bin $i$ and
$n_i$ no. of observations in bin $i$

:::

::: {.column width="50%"}
### Logistic
\begin{equation*}
P(u) = a \frac{1+ m \exp(-u/\tau)}{1+ n \exp(-u/\tau)}
\end{equation*}
where $a$ represents the upper asymptote, $n, m$ shape the lower asymptote, and $\tau$ controls the transition. 
:::

:::

## Power curve modelling

::: {.columns}

::: {.column width="50%"}

### 5 parameter curve
\begin{equation*}
  P(u) = D+ \frac{A-D}{(1+(u/C)^B)^G} 
\end{equation*}
where: $A$ and $D$ are the upper and lower asymptotes, $C$ is the inflection point, $B$ the slope at inflection point, adn $G$ controls the asymmetry.

:::

::: {.column width="50%"}

:::

:::

## Reanalysis data to quantify extreme wind power statistics
::: {.columns}

::: {.column width="50%"}
\footnotesize

D. Cannon, D. Brayshaw, et. al (2015)

- MERRA wind speed validated with MIDAS
- Vertical interpolation with a logarithmic change
- Calibrated power curves based on manufacturers PC
- Use that to analys extreme low and high levels, and ramps
:::

::: {.column width="50%"}
\centering
\small $\Delta$CF comparison 
\includegraphics[width=\linewidth]{../fig_extra/2013_cannon_delta_comparison.png}
:::

:::

## Balancing energy
::: {.columns}

::: {.column width="50%"}
\footnotesize
H. Thornton, D. Brayshaw analyse the relationship between weather, energy demand, and wind power.

- ERA Interim wind speed
- Cubic power curve with air density correction
- CF compared with GB average from other studies
- Seasonal effects on Demand and 

:::

::: {.column width="50%"}
\centering
\small $\Delta$CF comparison 
\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig_extra/2020_thornton_CF_vs_demand.png}
:::

:::

## Analysis of extreme wind droughts
::: {.columns}

::: {.column width="50%"}
\footnotesize
Panit Potisomporn, C. Vogel (2024) perform a extreme value analysis of wind droughts in GB.

- Use ERA5 wind speeds calibrated to MIDAS with QM
- Build a ML algorithm that learns how to extrapolate wind speed from 10m to hub height
- Use a 5 parameter logistic function to model power curve
- Model energy losses with factors by type.

:::

::: {.column width="50%"}
\centering
\small CF calibration 
\includegraphics[width=\linewidth]{../fig_extra/2024_potisomporn_CF_comparison.png}
:::

:::


## Next Steps

\begin{itemize}
   \item Power curve parametric model from data
   \item Check for spatial correlation 
   \item Quantile mapping calibration
\end{itemize}


<!-- 
# Appendix

## ERA5 data {#sec-era5}

\begin{columns}[T, totalwidth=\textwidth]
   \begin{column}{0.4\textwidth}
    % set font size for itemize levels locally
    {
    \setbeamerfont{itemize/enumerate body}{size=\fontsize{9}{10}\selectfont}
    \setbeamerfont{itemize/enumerate subbody}{size=\fontsize{8}{10}\selectfont}

    \begin{itemize}
      \item 5th generation of ECMWF's global reanalysis
      \item Spatial resolution $0.25^\circ \times 0.25^\circ$ (31km $\times$ 31 km at equator)
      \item Hourly temporal resolution
      \item Heights: 10m, 100m
    \end{itemize}
    } % end local font block
  \end{column}

  % Right column: figure
  
  \begin{column}{0.6\textwidth}
    \centering
    \small Average monthly wind speed and 90 CI
    \includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/ws_uk_1s.pdf}
  \end{column}

\end{columns}

## ERA5 series

::: {.columns}

::: {.column width="50%"}
\centering
    \small Wind speed seasonality
    \includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/ws_uk_1s_seas.eps}
:::

::: {.column width="50%"}
\centering
    \small Wind speed by location type
    \includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/ws_uk_tech.eps}
:::

:::


## ERA5 wind direction at wind farms

::: {.columns}

::: {.column width="50%"}
\centering
    \small Wind direction frequency
    \includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/wd_uk_1s.eps}
:::

::: {.column width="70%"}
\centering
    \small Seasonal patterns
    \includegraphics[width=\linewidth,height=7cm,keepaspectratio]{../fig/wd_uk_1s_seas.eps}
:::

:::


# Elexon data {-}

## BMU wind data

\begin{columns}[T, totalwidth=\textwidth]
   \begin{column}{0.5\textwidth}
    % set font size for itemize levels locally
    {
    \setbeamerfont{itemize/enumerate body}{size=\fontsize{9}{10}\selectfont}
    \setbeamerfont{itemize/enumerate subbody}{size=\fontsize{8}{10}\selectfont}

    \begin{itemize}
      \item 150 wind farms split in over 216 units
      \item Total capacity: 27 GW
      \item Half hourly resolution
      \item Records starting in 2019
      \item Curtailment and outage data available
      \item Location / turbine data unavailable
    \end{itemize}
    } % end local font block
  \end{column}

  % Right column: figure
  \begin{column}{0.5\textwidth}
    \centering
    \small Wind installed capacity
    \includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/wind_capacity_type_2025_nt.eps}
  \end{column}

\end{columns}

## BMU wind data

\small BMU wind generation (2024) 
```{r}
pot_summary <- fread(
  file.path("../data/hist_pot.csv")
)

# on shore off shore distribution
pot_summary %>% 
    group_by(tech_typ) %>%
    summarise(
        total_gen = sum(quantity, na.rm = TRUE)/1000,
        total_curt = sum(curtailment, na.rm = TRUE)/1000,
        total_pot = sum(potential, na.rm = TRUE)/1000,
        total_cap = sum(capacity, na.rm = TRUE),
        n_hours = sum(n_hours),
        n = n()
    ) %>% 
    janitor::adorn_totals("row") %>%
    mutate(
        cap_factor = total_gen*1000 / (total_cap )/(n_hours/n/2),
        pot_factor = total_pot*1000 / (total_cap )/(n_hours/n/2)
    ) %>% select(-n_hours) %>% 
    kable(
        col.names = c("Type", "Generation (GWh)", "Curtailment", "Potential" ,"Capacity (MW)", "Number of BMUs", "CF","PF"),
        digits = c(0,0,0,0,0,0,2,2),
        format = "latex",
        format.args = list(big.mark = ",")) %>%
    kable_styling(latex_options = c("scale_down"))
```

## Capacity factor through time

::: {.columns}

::: {.column width="50%"}
\centering
    \small Capacity factor by year
    \includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/capacity_factor2.pdf}
:::

::: {.column width="50%"}
\centering
    \small Capacity factor by quarter
    \includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/capacity_factor2_q.pdf}
:::

:::

## Curtailment through time

::: {.columns}

::: {.column width="50%"}
\centering
    \small Curtailment by year
    \includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/curtailment_factor.pdf}
:::

::: {.column width="50%"}
\centering
    \small Curtailment seasonality
    \includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/curtailment_factor_q.pdf}
:::

:::

## Curtailment

::: {.columns}

::: {.column width="50%"}
\centering
    \footnotesize Curtailment map
    \includegraphics[width=\linewidth,height=6cm,keepaspectratio]{../fig/curtailment_map.pdf}
:::

::: {.column width="50%"}
\centering
 \footnotesize Curtailment by region

![](../fig/curtailment_monthly.pdf){width=100% height=2.5cm}

\footnotesize Average during curtailment periods

![](../fig/curtailment_plus_monthly.pdf){width=100% height=2.5cm}


:::

:::


## Curtailment through the day

::: {.columns}

::: {.column width="50%"}
\centering
    \small Curtailment by hour
    \includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/curtailment_intraday.pdf}
:::

::: {.column width="50%"}
\centering
    \small Average during curtailment periods
    \includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/curtailment_plus_intraday.pdf}
:::

:::

## Potential generation

::: {.columns}

::: {.column width="50%"}
\centering
    \small Potential by year
    \includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/potential_factor.pdf}
:::

::: {.column width="50%"}
\centering
    \small Potential seasonality
    \includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/potential_factor_q.pdf}
:::

:::

## Renewable energy planning database (REPD)

```{r}
repd0 <- fread(file.path(path,"renew_energ_plan_db", "repd-q2-jul-2025.csv")) %>% 
    rename(
    tech_typ = `Technology Type`,
    devel_stat = `Development Status (short)`,
    capacity_repd = `Installed Capacity (MWelec)`,
    capacity_turb = `Turbine Capacity`,
    n_turb = `No. of Turbines`,
    height_turb = `Height of Turbines (m)`,
    x_coord = `X-coordinate`,
    y_coord = `Y-coordinate`)
names(repd0) <- tolower(gsub(" ", "_", names(repd0)))

```

::: {.columns}

::: {.column width="50%"}

\setbeamerfont{itemize/enumerate body}{size=\fontsize{9}{10}\selectfont}
\setbeamerfont{itemize/enumerate subbody}{size=\fontsize{8}{10}\selectfont}
\begin{itemize}
      \item Official UK government renewable data
      \item Over 800 wind farms listed as operational
      \item Coordinates available 
      \item Also available:
      \begin{itemize}
        \item Development status 
        \item Number of turbines
        \item Turbine capacity
        \item Turbine height (for some only)
      \end{itemize}
    \end{itemize}
:::

::: {.column width="50%"}
\centering
\small REPD wind farms
\vspace{0.2cm}
```{r}
reduced_table <- repd0 %>%
  filter(
    grepl("Wind", tech_typ),
    devel_stat %in% c("Operational", "Under Construction")
  ) %>%
  group_by(tech_typ, devel_stat) %>%
  summarise(
    n = n_distinct(ref_id),
    capacity = sum(capacity_repd, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  janitor::adorn_totals(where = "row")

kable(
  reduced_table,
  col.names = c("Type", "Development Status","Count","Capacity MW"),
  digits = 0,
  format = "latex",
  format.args = list(big.mark = ",")
) %>% 
    kable_styling(
        latex_options = c("scale_down")
    )
```
:::
:::

## Turbine data available

::: {.columns}

::: {.column width="50%"}
\centering
    \small Turbine capacity histogram
    \includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/hist_turb_capacity_2025.eps}
:::

::: {.column width="50%"}
\centering
    \small Number of turbines histogram
    \includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/hist_n_turb_2025.eps}
:::

:::

## Turbine data available

::: {.columns}

::: {.column width="50%"}
\centering
Turbine Height availability
\vspace{1cm}
```{r}
ref_catalog_2025 <- fread(
  file.path("../data/ref_catalog_wind_2025.csv.gz")) 
ref_catalog_2025 %>% 
    mutate(heightNA = is.na(height_turb)) %>% 
    group_by(tech_typ, heightNA) %>% 
    summarise(
        height = mean(height_turb, na.rm = TRUE),
        n = n()
    ) %>% 
    kable(
        col.names = c("Type","Height available", "Average height (m)","count"),
        digits = 1,
        format = "latex"
    ) %>% 
    kable_styling(
        latex_options = c("scale_down")
    )

```

:::

::: {.column width="50%"}
\centering
    \small Turbine height histogram
    \includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/hist_turb_height_2025.eps}
:::

:::

# Outages {-}
## Outage data
::: {.columns}

::: {.column width="50%"}
\begin{itemize}
  \item REMIT messages announce plan or unplanned outages
  \item event start / end times
  \item affected BMUs and capacity reduction
\end{itemize}
:::

::: {.column width="50%"}
\centering
    \small Duration of events
    \includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/remit_events_by_year.pdf}
:::

:::
## Outage characteristics
::: {.columns}

::: {.column width="50%"}
\centering
    \small Number of events by year
    \includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/remit_event_duration.pdf}
:::

::: {.column width="50%"}
\centering
    \small Duration of events
    \includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/remit_capacity_impact_perc.eps}
:::

:::

## Outages characteristics by location
::: {.columns}

::: {.column width="32%"}
\centering
    \small No. of events
    \includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/remit_nevents_map.pdf}
:::

::: {.column width="32%"}
\centering
    \small Mean duration
    \includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/remit_duration_map.pdf}
:::

::: {.column width="32%"}
\centering
    \small Unavailable capacity
    \includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/remit_cap_imp_map.pdf}
:::

:::

# Power conversion {-}

## Generic power curves

::: {.columns}

::: {.column width="50%"}

\setbeamerfont{itemize/enumerate body}{size=\fontsize{9}{10}\selectfont}
\setbeamerfont{itemize/enumerate subbody}{size=\fontsize{8}{10}\selectfont}
\begin{itemize}
      \item Using 3 generic power curves
      \item Offshore plus the IEC 3 classes
      \item Assigning class based on GWA mean wind speed at location
      \item Rescaling rated power to turbine capacity
\end{itemize}

\vspace{0.2cm}
\centering
IEC classification

\vspace{0.2cm}
\resizebox{\linewidth}{!}{%
\begin{tabular}{l c c l}
\toprule
Class & Mean wind speed  & Extreme 10-min & Typical sites \\
 & at hub height (m/s) &  gust (m/s) \\
\midrule
I   & 10   & 70   & Very windy / exposed sites \\
II  & 8.5  & 59.5 & Moderate wind sites \\
III & 7.5  & 52.5 & Low-wind / inland sites \\
\bottomrule
\end{tabular}
}
:::

::: {.column width="50%"}
\centering
    \small Generic power curves (PC$_k$)
    \includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/generic_power_curves_uncens.eps}
:::

:::
## Power estimate based on generic power curves

For each location $i$ we have: 

- observed raw power $\tilde p_{it}$
- curtailment amount $a_{it}$
- potential output $p_{it} = \tilde p_{it} + a_{it}$
- wind farm capacity $c_i$
- turbine height $v_i$
- ERA5 wind speed $w_{it}$.

## Power estimate based on generic power curves

An initial estimate will require:
\begin{itemize}
  \item Mapping location to a rescaled power curve $\widetilde \text{PC}_k$
  \item Estimate wind farm power in GWh
\end{itemize}

\begin{equation*}
    \hat p_{it} = \widetilde \text{PC}_{k} (w_{it})
\end{equation*}

## ERA5 based estimates vs Elexon 2024

\begin{enumerate}
  \item Remove periods with outages
  \begin{itemize}
    \item Modelling loss due to expected outages could be future work
  \end{itemize}
  \item Vertical interpolation of ERA5 wind speed to turbine height
  \item Apply generic power curve to get estimated power
  \begin{itemize}
    \item Modelling power curve from data could be future work
  \end{itemize}
\end{enumerate}

# Curtailment and outages in select wind farms {-}

## Generic power curves vs observed data - Fallago Rig
Onshore wind farm with moderate curtailment
\begin{columns}[T, totalwidth=\textwidth]

  \begin{column}{0.5\textwidth}
    \centering
    {\small Comparison with raw generation}
    \includegraphics[
      width=\linewidth,
      height=5cm,keepaspectratio]{
        ../fig/pc_comparison_T_FALGW-1_valq0.pdf}
    
  \end{column}

  \begin{column}{0.5\textwidth}
    \centering
    {\small Adding curtailment}

    \includegraphics[
      width=\linewidth,
      height=5cm,keepaspectratio]{
        ../fig/pc_comparison_T_FALGW-1_valpot.pdf}
  \end{column}

\end{columns}

## Generic power curves vs observed data - Fallago Rig
Onshore wind farm with moderate curtailment
\begin{columns}[T, totalwidth=\textwidth]

  \begin{column}{0.5\textwidth}
    \centering
    {\small Highlighting outages}
    \includegraphics[
      width=\linewidth,
      height=5cm,keepaspectratio]{
        ../fig/pc_comparison_T_FALGW-1_valout.pdf}
    
  \end{column}

  \begin{column}{0.5\textwidth}
    \centering
    {\small Potential generation excluding outages}

    \includegraphics[
      width=\linewidth,
      height=5cm,keepaspectratio]{
        ../fig/pc_comparison_T_FALGW-1_valnout.pdf}
  \end{column}

\end{columns}

## Generic power curves vs observed data - Viking
Onshore wind farm with high curtailment
\begin{columns}[T, totalwidth=\textwidth]

  \begin{column}{0.5\textwidth}
    \centering
    {\small Comparison with raw generation}
    \includegraphics[
      width=\linewidth,
      height=5cm,keepaspectratio]{
        ../fig/pc_comparison_T_VKNGW-1_valq0.pdf}
    
  \end{column}

  \begin{column}{0.5\textwidth}
    \centering
    {\small Adding curtailment}

    \includegraphics[
      width=\linewidth,
      height=5cm,keepaspectratio]{
        ../fig/pc_comparison_T_VKNGW-1_valpot.pdf}
  \end{column}

\end{columns}

## Generic power curves vs observed data - Viking
Onshore wind farm with high curtailment
\begin{columns}[T, totalwidth=\textwidth]

  \begin{column}{0.5\textwidth}
    \centering
    {\small Highlighting outages}
    \includegraphics[
      width=\linewidth,
      height=5cm,keepaspectratio]{
        ../fig/pc_comparison_T_VKNGW-1_valout.pdf}
    
  \end{column}

  \begin{column}{0.5\textwidth}
    \centering
    {\small Potential generation excluding outages}

    \includegraphics[
      width=\linewidth,
      height=5cm,keepaspectratio]{
        ../fig/pc_comparison_T_VKNGW-1_valnout.pdf}
  \end{column}

\end{columns}

## Generic power curves vs observed data - Hornsea 1
Offshore wind farm with low curtailment
\begin{columns}[T, totalwidth=\textwidth]

  \begin{column}{0.5\textwidth}
    \centering
    {\small Comparison with raw generation}
    \includegraphics[
      width=\linewidth,
      height=5cm,keepaspectratio]{
        ../fig/pc_comparison_HOWAO-1_valq0.pdf}
    
  \end{column}

  \begin{column}{0.5\textwidth}
    \centering
    {\small Adding curtailment}

    \includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/pc_comparison_HOWAO-1_valpot.pdf}
  \end{column}

\end{columns}

## Generic power curves vs observed data - Hornsea 1
Offshore wind farm with low curtailment
\begin{columns}[T, totalwidth=\textwidth]

  \begin{column}{0.5\textwidth}
    \centering
    {\small Highlighting outages}
    \includegraphics[
      width=\linewidth,
      height=5cm,keepaspectratio]{
        ../fig/pc_comparison_HOWAO-1_valout.pdf}
    
  \end{column}

  \begin{column}{0.5\textwidth}
    \centering
    {\small Potential generation excluding outages}

    \includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/pc_comparison_HOWAO-1_valnout.pdf}
  \end{column}

\end{columns}

## Generic power curves vs observed data - Seagreen 1
Offshore wind farm with high curtailment
\begin{columns}[T, totalwidth=\textwidth]

  \begin{column}{0.5\textwidth}
    \centering
    {\small Comparison with raw generation}
    \includegraphics[
      width=\linewidth,
      height=5cm,keepaspectratio]{
        ../fig/pc_comparison_T_SGRWO-1_valq0.pdf}
    
  \end{column}

  \begin{column}{0.5\textwidth}
    \centering
    {\small Adding curtailment}

    \includegraphics[
      width=\linewidth,
      height=5cm,keepaspectratio]{
        ../fig/pc_comparison_T_SGRWO-1_valpot.pdf}
  \end{column}

\end{columns}

## Generic power curves vs observed data - Seagreen 1
Offshore wind farm with high curtailment
\begin{columns}[T, totalwidth=\textwidth]

  \begin{column}{0.5\textwidth}
    \centering
    {\small Highlighting outages}
    \includegraphics[
      width=\linewidth,
      height=5cm,keepaspectratio]{
        ../fig/pc_comparison_T_SGRWO-1_valout.pdf}
    
  \end{column}

  \begin{column}{0.5\textwidth}
    \centering
    {\small Potential generation excluding outages}

    \includegraphics[
      width=\linewidth,
      height=5cm,keepaspectratio]{
        ../fig/pc_comparison_T_SGRWO-1_valnout.pdf}
  \end{column}

\end{columns} -->