---
title: "ERA 5 calibration to Elexon power"
author:
  - "S. Gomez$^1$"
institute:
   - $^1$School of Mathematics, University of Edinburgh
institute-short: "UoE"
date: "2026-07-14"
format: 
  beamer:
    theme: Rochester
    colortheme: spruce
    fonttheme: professionalfonts
    fig-cap-location: top
    toc: true
    slide-level: 2
    split-bib: true
    output-file: "wpow_calibration_slides_26-07-14.pdf"
header-includes:
  - \usepackage{graphicx}
  - \setbeamerfont{itemize/enumerate body}{size=\footnotesize}
  - \setbeamerfont{itemize/enumerate subbody}{size=\footnotesize}
  - \setbeamerfont{itemize/enumerate subsubbody}{size=\footnotesize}
  - \usepackage{tikz}
  - \usetikzlibrary{positioning, fit, arrows.meta, decorations.pathreplacing, backgrounds}
fontsize: 12pt
aspectratio: 169
editor: source
bibliography: calib_pow_ref.bib
nocite: "@*"
---

```{r include=FALSE}
require(tidyverse)
require(kableExtra)
require(parallel)
require(dplyr)
require(scales)
require(ggsci)

require(jsonlite)
require(httr)

require(data.table)
require(sf)
library(rnaturalearth)
library(rnaturalearthdata)

require(plotly)
require(ggthemes)
require(arrow)
path <- "~/Documents/elexon"

```

```{r}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo = FALSE, cache = FALSE)
theme_set(theme_bw())
```

## Updates {#sec-updates .unnumbered .nolink}

### New
\begin{itemize}
  \item Updated validation framework
  \item OOS assessment
  \begin{itemize}
    \item OOS time
    \item OOS locations
  \end{itemize}
  \item Low wind events distribution
\end{itemize}




# Model validation framework

## Updated models
\begin{itemize}
    \item Wind farm level models
    \begin{enumerate}    
        \item Linear model (LM)
        \item Linear model with hour effect
        \item Linear model with AR1 (or AR2) errors
        \item Spatio-temporal model
        \item Quantile mapping (QM)
   \end{enumerate}
   \item Models on Great Britain aggregated power
        \begin{enumerate}[6]
            \item Linear model (GB LM)
        \end{enumerate}
\end{itemize}

## Model equation
### Linear model (LM)

```{=latex}
        \begin{align}
        \mathbb{E}(p^{\mathrm{calib}}_{it})
        &= \alpha_{k} + \beta_{k} \hat{p}_{it} + \gamma e_i + \delta d_i + s_{w,k}(w_{it})
        + \epsilon_{it},\label{eq:base_model}\\
        \epsilon_{it} & \sim \mathcal{N}(0, \sigma_\epsilon) \nonumber
        \end{align}
        where the intercept and slope vary by type $k:\{\text{onshore, offshore}\}$, $e_i$ and $d_i$ represent the land elevation and the distance to the coast at location $i$, respectively, and $s_{w,k}(\cdot)$ is a smooth function that serves as a power regime correction.
```
## Alternative models
### Spatiotemporal model (ST)

```{=latex}
\begin{align}
\mathbb{E}(p^{\mathrm{calib}}_{it})
&= \alpha_{k} + \beta_{k} \hat{p}_{it} + \gamma e_i + \delta d_i + s_{w,k}(w_{it}) + u_t,\tag{4}\label{eq:spt_model}\\
u_{it} &= \xi_i \cdot \tau_t.\nonumber
\end{align}
The additional term $u_{it}$ is a separable spatiotemporal random effect. It consists of $\xi_i$, a zero-mean Gaussian spatial field over farm locations, and $\tau_t$, a temporal component modelled as an AR(1) process.
```


## Framework for OOS assessment

::: {.columns}

::: {.column width="50%"}
<!-- \footnotesize -->
- Run models on a sample of days
  - 80/20 WF split for training and testing
  - Training with 3 days to predict the 4th
  - Randomly chosen 15 days from 2025, 5 per regime

\vspace{0.5cm}
\centering
\begin{tikzpicture}[
  scale=0.7, transform shape,
    box/.style={draw, minimum width=2cm, minimum height=1.2cm, align=center, font=\small},
    train/.style={box, fill=blue!15},
    pred/.style={box, fill=orange!25},
    lbl/.style={font=\scriptsize, align=center}
]

    % Training days
    \node[train] (d1) at (0,0) {Day $t-2$};
    \node[train] (d2) at (2.2,0) {Day $t-1$};
    \node[train] (d3) at (4.4,0) {Day $t$};
    
    % Prediction day
    \node[pred] (d4) at (7.2,0) {Day $t+1$};
    
    % Arrow from training block to prediction
    \draw[-{Latex[length=3mm]}] (3.4,-1) -- (6.2,-1);
    
    % Labels underneath
    \node[lbl, below=0.3cm of d1] {};
    \node[lbl] at (2.2,-0.9) {Training data};
    \node[lbl] at (7.2,-0.9) {Prediction};
    
    % Bracket over training days
    \draw[decorate, decoration={brace, amplitude=6pt}] (-1,0.8) -- (5.4,0.8)
        node[midway, above=6pt, font=\small] {3 days of training};

\end{tikzpicture}


:::

::: {.column width="50%"}
\centering
\small Sampled days for performance assessment

\includegraphics[width=\linewidth,height=6cm,keepaspectratio]{../fig/batch2025/trainsplit_wind_farm_locations_v0.pdf}
:::

:::


## Sampled days in different regimes

::: {.columns}

::: {.column width="50%"}
- Objective is to assess performance by regime
- t+1 day are from specific regimes of wind speed
- Randomly chosen 5 days per regime
- Previous 3 days can be anything
:::

::: {.column width="50%"}
\centering
\small Sampled days for performance assessment

\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/batch2025/daily_generation_time_series_sampled_days.pdf}
:::

:::

## Key results

::: {.columns}

::: {.column width="50%"}
- Increasing model complexity improves mean power estimates
- However, uncertainty quantification for complex models misreprestents observed variability

\tiny

\begin{table}[!h]
\centering
\begin{tabular}{lcc}
\toprule
Model & RMSE (IS) & RMSE (OOS) \\
\midrule
GB LM                  & 0.212 & 0.203 \\
AR1 model              & 0.184 & 0.193 \\
AR2 model              & 0.184 & 0.194 \\
Linear model           & 0.189 & 0.188 \\
LM+hour model          & 0.061 & 0.193 \\
\addlinespace
Spatio-temporal coarse & 0.104 & 0.182 \\
Spatio-temporal fine   & 0.071 & 0.171 \\
\bottomrule
\end{tabular}
\end{table}

:::

::: {.column width="50%"}
\centering
\small Coverage for day-ahead prediction bands

\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/batch2025/WF_pred_band_coverage.pdf}
:::

:::

# Out-of-sample analysis


## Wind farm prediction bands (oos time)

::: {.columns}

::: {.column width="50%"}
\centering
\small Linear model (27-01-2025)

\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/batch2025/WF_pred_band_lm_250127_2.pdf}
:::

::: {.column width="50%"}
\centering
\small Spatiotemporal model coarse (27-01-2025)

\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/batch2025/WF_pred_band_st0_m1_250127_2.pdf}
:::

:::

## Wind farm prediction bands (oos time)

::: {.columns}

::: {.column width="50%"}
\centering
\small LM+hour model (27-01-2025)

\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/batch2025/WF_pred_band_spde1d_250127_2.pdf}
:::

::: {.column width="50%"}
\centering
\small AR1 model (27-01-2025)

\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/batch2025/WF_pred_band_ar1_250127_2.pdf}
:::

:::

## Aggregated prediction bands (oos time)

::: {.columns}

::: {.column width="50%"}
\centering
\small GB level

\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/batch2025/GB_pred_band_250127.pdf}
:::

::: {.column width="50%"}
\centering
\small Scatter plot oos

\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/batch2025/GB_fit_scatter_oos.pdf}
:::

:::


## Day-ahead forecast coverage (oos time)

::: {.columns}

::: {.column width="50%"}
- Models at WF can achieve target coverage
- Aggregation reduces coverage
- High regime more succeptible to coverage loss
- Spatiotemporal model is not improving coverage
:::

::: {.column width="50%"}
\centering
\small Coverage for day-ahead predictions

\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/batch2025/WF_pred_band_coverage.pdf}
:::

:::


## Coverage under aggregation (oos time)

::: {.columns}

::: {.column width="50%"}
\centering
\small GB level

\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/batch2025/GB_pred_band_coverage.pdf}
:::

::: {.column width="50%"}
\centering
\small Onshore / offshore

\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/batch2025/GB_pred_band_coverage_tech.pdf}
:::

:::

## Coverage under aggregation by regime (oos time)
\centering
\small Regime aggregation

\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/batch2025/GB_pred_band_coverage_regime.pdf}



# Out-of-sample locations {-}

## Wind farm prediction bands (oos space)

::: {.columns}

::: {.column width="50%"}
\centering
\small Linear model (27-01-2025)

\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/batch2025/WF_pred_band_spaceoos_lm_250127_1.pdf}
:::

::: {.column width="50%"}
\centering
\small Spatiotemporal model coarse (27-01-2025)

\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/batch2025/WF_pred_band_spaceoos_st0_m1_250127_1.pdf}
:::

:::

## Wind farm prediction bands (oos space)

::: {.columns}

::: {.column width="50%"}
\centering
\small LM+hour model (27-01-2025)

\includegraphics[width=\linewidth,height=6cm,keepaspectratio]{../fig/batch2025/WF_pred_band_spaceoos_spde1d_250127_1.pdf}
:::

::: {.column width="50%"}
\centering
\small AR1 model (27-01-2025)

\includegraphics[width=\linewidth,height=6cm,keepaspectratio]{../fig/batch2025/WF_pred_band_spaceoos_ar1_250127_1.pdf}
:::

:::

## Aggregated prediction bands (oos space)

::: {.columns}

::: {.column width="50%"}
\centering
\small GB level (27-01-2025)

\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/batch2025/GB_pred_band_spaceoos_250127.pdf}
:::

::: {.column width="50%"}
\centering
\small Scatter plot oos

\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/batch2025/GB_fit_scatter_oos_space.pdf}
:::

:::


## Unseen locations prediction intervals coverage (oos space)

::: {.columns}

::: {.column width="50%"}
- Attaining coverage in unseen locations is more challenging
- Isolated locations can lead to high uncertainty
- Location specific events may affect coverage
- Spatiotemporal model is not improving coverage
:::

::: {.column width="50%"}
\centering
\small Coverage for new locations

\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/batch2025/WF_pred_band_coverage_spaceoos.pdf}
:::

:::


## Coverage under aggregation (oos space)

::: {.columns}

::: {.column width="50%"}
\centering
\small GB level

\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/batch2025/GB_pred_band_coverage_spaceoos.pdf}
:::

::: {.column width="50%"}
\centering
\small Onshore / offshore

\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/batch2025/GB_pred_band_coverage_tech_spaceoos.pdf}
:::

:::

## Coverage under aggregation by regime (oos space)
\centering
\small Regime aggregation

\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/batch2025/GB_pred_band_coverage_regime_spaceoos.pdf}

# Low wind speed events distribution

## Low wind events (LWE) analisys

- Calibration should also consider LWE
- Using 0.1 as threshold for LWE
- Testing that model matches:
  - observed frequency of events
  - distribution of duration

## Low wind speed events (insample)

::: {.columns}

::: {.column width="50%"}
\centering
Low wind speed events distribution
\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/batch2025/low_wind_duration_dist_t0_1.pdf}
:::

::: {.column width="50%"}
\centering
Low wind speed events distribution
\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/batch2025/low_wind_duration_qq_t0_1.pdf}
:::

:::

## Low wind speed events (oos time)

::: {.columns}

::: {.column width="50%"}
\centering
Low wind speed events distribution
\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/batch2025/pred_samples_lwe_duration_dist_t0_1.pdf}
:::

::: {.column width="50%"}
\centering
Low wind speed events distribution
\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/batch2025/pred_samples_lwe_duration_qq_t0_1.pdf}
:::

:::

## Low wind speed events (oos space)

::: {.columns}

::: {.column width="50%"}
\centering
Low wind speed events distribution
\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/batch2025/newloc_samples_lwe_duration_dist_t0_1.pdf}
:::

::: {.column width="50%"}
\centering
Low wind speed events distribution
\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/batch2025/newloc_samples_lwe_duration_qq_t0_1.pdf}
:::

:::


# Next steps

## Addressing challenges

- Collapsing uncertainty under aggregation
- Spatiotemporal model is not improving calibration
- Low wind speed events are not well represented in the model samples

# Appendix {-}

## Alternative models
### Linear model with hour effect

```{=latex}

        This model results from adding a periodic hour effect $h_t$ to control temporal effects
        \begin{equation}
        \mathbb{E}(p^{\mathrm{calib}}_{it})
        = \alpha_{k} + \beta_{k} \hat{p}_{it} + \gamma e_i + \delta d_i + s_{w,k}(w_{it}) + h_t
        + \epsilon_{it},\label{eq:1dspde_model}
        \end{equation}
```
## Alternative models
### Linear model with AR1 (or AR2) errors

```{=latex}
        A more robust temporal correlation term can be added to the model:
        \begin{align}
        \mathbb{E}(p^{\mathrm{calib}}_{it})
        &= \alpha_{k} + \beta_{k} \hat{p}_{it} + \gamma e_i + \delta d_i + s_{w,k}(w_{it}) + u_t,\label{eq:ar_model}\\
        u_t &\sim \mathrm{AR}(1) ,\nonumber
        \end{align}
        where $u_t$ captures residual temporal autocorrelation, satisfying $u_t = \rho\, u_{t-1} + \varepsilon_{it}$ with $|\rho| < 1$
```

## Alternative models
### Quantile mapping (QM)

```{=latex}
This approach doesn't include covariates but instead of correcting only the mean bias, makes the model match the distribution of observed power.
        \begin{equation}
        p^{\mathrm{calib}}_{t_k}
        = F_{\mathrm{obs}}\!\left(F_{\mathrm{mod}}^{-1}(\hat{p}_{t_k})\right),
        \tag{5}
        % \tag{\ref{eq:qm_model}}
        \end{equation}
        where $F_{\mathrm{mod}}$ is the empirical CDF of the ERA5-derived power estimate and $F_{\mathrm{obs}}$ is the empirical CDF of the observed generation. The transfer function $F_{\mathrm{obs}} \circ F_{\mathrm{mod}}^{-1}$ is a non-decreasing smooth function that corrects distributional bias by power regime.
```
## Alternative models
### Linear model (GB LM)

```{=latex}

            If the goal is only to produce estimates at the aggregate level, the optimal correction is achieved by modelling aggregate time series instead of aggregating estimates.
            \begin{align}
            p_t &= \sum_i p_{it} \nonumber\\
            \mathbb{E}(p^{\mathrm{calib}}_{t})
            &= \alpha_{k} + \beta_{k} \hat{p}_{t} + s_{w,k}(\tilde w_{t})
            + \epsilon_{t},\label{eq:agg_model}\\
            \epsilon_{t} & \sim \mathcal{N}(0,\tilde \sigma_{\epsilon}), \nonumber \tag{6}\label{eq:agg_error}
            \end{align}
            where now wind power $p_t$ is the average capacity factor of all the wind farms, $\tilde w_t$ is the average wind speed of the observed locations, and $\epsilon$ now represents the error in aggregated terms.
```
        

## Meshes for spatiotemporal models

::: {.columns}

::: {.column width="50%"}
\centering
Coarse mesh

\includegraphics[width=\linewidth,height=6cm,keepaspectratio]{/home/s2441782/Documents/elexon/extra/mesh_figs/spatial_mesh_coarse_250121.pdf}
:::

::: {.column width="50%"}
\centering
Fine mesh

\includegraphics[width=\linewidth,height=6cm,keepaspectratio]{/home/s2441782/Documents/elexon/extra/mesh_figs/spatial_mesh_fine_250121.pdf}
:::

:::

# Error metrics and performance assessment {-}
## Error metrics

\footnotesize
\begin{table}[!h]
\centering
\begin{tabular}[t]{lcccccccc}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{2}{c}{RMSE} & \multicolumn{2}{c}{MAE} & \multicolumn{2}{c}{MDAPE (\%)} & \multicolumn{2}{c}{Bias} \\
\cmidrule(l{3pt}r{3pt}){2-3} \cmidrule(l{3pt}r{3pt}){4-5} \cmidrule(l{3pt}r{3pt}){6-7} \cmidrule(l{3pt}r{3pt}){8-9}
Model & IS & OOS & IS & OOS & IS & OOS & IS & OOS\\
\midrule
GB LM & 0.212 & 0.203 & 0.147 & 0.138 & 31.8 & 36.7 & 0.024 & 0.047\\
AR1 model & 0.184 & 0.193 & 0.135 & 0.137 & 30.4 & 37.2 & 0.010 & 0.055\\
AR2 model & 0.184 & 0.194 & 0.135 & 0.137 & 30.4 & 37.3 & 0.010 & 0.056\\
Linear model & 0.189 & 0.188 & 0.132 & 0.130 & 31.9 & 36.5 & 0.001 & 0.033\\
LM+hour model & 0.061 & 0.193 & 0.041 & 0.134 & 9.8 & 41.7 & 0.004 & 0.041\\
\addlinespace
Spatio-temporal coarse & 0.104 & 0.182 & 0.070 & 0.124 & 16.5 & 40.2 & 0.005 & 0.032\\
Spatio-temporal fine & 0.071 & 0.171 & 0.046 & 0.113 & 10.0 & 37.2 & 0.003 & 0.042\\
\bottomrule
\end{tabular}
\end{table}

## Error metrics

```{r}
  
aggr_summary_tbl <- read.csv("../summaries/calib_metrics_batch2025.csv")
  
aggr_summary_tbl%>%
  mutate(Bias = formatC(Bias, format = "e", digits = 2)) %>% 
  kable(digits = c(3,3,3,3), format = "latex") %>% 
  kable_styling(
    latex_options = c("scale_down")
  )
```

## Error metrics by regime
\vspace{-0.5cm}
```{r regimesummary}
  
aggr_summary_tbl <- read.csv("../summaries/calib_metrics_batch2025_regime.csv")
  
aggr_summary_tbl%>%
  mutate(
    across(matches("Bias"), ~ formatC(., format = "e", digits = 2))
    ) %>%  
  kbl(
  format = "latex",
  col.names = c("Model", rep( c("RMSE", "Bias","MAE"), 3)),
  digits = 3,
  booktabs = TRUE,
  align = c("l", rep("c", 9))
) %>%
  add_header_above(c(
    " " = 1,
    "Low" = 3,
    "Mid" = 3,
    "High" = 3
  )) %>% 
  kable_styling(
    latex_options = c("scale_down")
  )
```

## Error metrics by technology
\vspace{-0.5cm}
```{r techsummary}

aggr_summary_tbl <- read.csv("../summaries/calib_metrics_batch2025_tech.csv")

aggr_summary_tbl%>%
  mutate(
    across(matches("Bias"), ~ formatC(., format = "e", digits = 2))
    # Bias = formatC(Bias, format = "e", digits = 2)
    ) %>% 
  kbl(
  format = "latex",
  col.names = c("Model", rep( c("RMSE", "Bias","MAE"), 2)),
  digits = 3,
  booktabs = TRUE,
  align = c("l", rep("c", 9))) %>%
  add_header_above(c(
    " " = 1,
    "Offshore" = 3,
    "Onshore" = 3
  )) %>% 
  kable_styling(
    latex_options = c("scale_down")
  )
```


## References

