---
title: "ERA 5 calibration to Elexon power"
author:
  - "S. Gomez$^1$"
institute:
   - $^1$School of Mathematics, University of Edinburgh
institute-short: "UoE"
date: "2026-07-09"
format: 
  beamer:
    theme: Rochester
    colortheme: spruce
    fonttheme: professionalfonts
    fig-cap-location: top
    toc: true
    slide-level: 2
    split-bib: true
    output-file: "wpow_calibration_slides_26-07-09.pdf"
header-includes:
  - \usepackage{graphicx}
  - \setbeamerfont{itemize/enumerate body}{size=\footnotesize}
  - \setbeamerfont{itemize/enumerate subbody}{size=\footnotesize}
  - \setbeamerfont{itemize/enumerate subsubbody}{size=\footnotesize}
fontsize: 12pt
aspectratio: 169
editor: source
bibliography: calib_pow_ref.bib
nocite: "@*"
---

```{r include=FALSE}
require(tidyverse)
require(kableExtra)
require(parallel)
require(dplyr)
require(scales)
require(ggsci)

require(jsonlite)
require(httr)

require(data.table)
require(sf)
library(rnaturalearth)
library(rnaturalearthdata)

require(plotly)
require(ggthemes)
require(arrow)
path <- "~/Documents/elexon"

```

```{r}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo = FALSE, cache = FALSE)
theme_set(theme_bw())
```

## Updates {#sec-updates .unnumbered .nolink}

### New
\begin{itemize}
  \item OOS assessment
  \begin{itemize}
    \item OOS time
    \item OOS locations
  \end{itemize}
  \item Low wind events distribution
\end{itemize}

# Model validation framework

## Framework for OOS assessment

::: {.columns}

::: {.column width="50%"}
<!-- \footnotesize -->
- Run models on a sample of days
  - 80/20 WF split for training and testing
  - Training with 3 days to predict the 4th
  - Randomly chosen 15 days from 2025, 5 per regime
- Evaluate performance at WF level
- Error under aggregation
  - GB level
  - Onshore / offshore

:::

::: {.column width="50%"}
\centering
\small Sampled days for performance assessment

\includegraphics[width=\linewidth,height=6cm,keepaspectratio]{../fig/batch2025/trainsplit_wind_farm_locations_v0.pdf}
:::

:::


## Sampled days in different regimes

::: {.columns}

::: {.column width="50%"}
- Objective is to assess performance by regime
- OOS days are different regimes of wind speed
- Previous 3 days can be anything
:::

::: {.column width="50%"}
\centering
\small Sampled days for performance assessment

\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/batch2025/daily_generation_time_series_sampled_days.pdf}
:::

:::

## Model portfolio


- Wind farm level models:
  - Generic Power Curve (ERA5 + generic PC estimate)
  - Quantile Mapping
  - Linear model 
  - AR1, AR2
  - Spatiotemporal model fine mesh and coarser mesh

- Linear model at Great Britain (GB) level

# Out-of-sample time


## Wind farm prediction bands

::: {.columns}

::: {.column width="50%"}
\centering
\small Linear model (27-01-2025)

\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/batch2025/WF_pred_band_lm_250127_2.pdf}
:::

::: {.column width="50%"}
\centering
\small Spatiotemporal model coarse (27-01-2025)

\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/batch2025/WF_pred_band_st0_m1_250127_2.pdf}
:::

:::

## Wind farm prediction bands

::: {.columns}

::: {.column width="50%"}
\centering
\small 1D SPDE model (27-01-2025)

\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/batch2025/WF_pred_band_spde1d_250127_2.pdf}
:::

::: {.column width="50%"}
\centering
\small AR1 model (27-01-2025)

\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/batch2025/WF_pred_band_ar1_250127_2.pdf}
:::

:::

## Aggregated prediction bands

::: {.columns}

::: {.column width="50%"}
\centering
\small GB level

\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/batch2025/GB_pred_band_250127.pdf}
:::

::: {.column width="50%"}
\centering
\small Scatter plot oos

\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/batch2025/GB_fit_scatter_oos.pdf}
:::

:::


## Day-ahead forecast coverage

::: {.columns}

::: {.column width="50%"}
- Models at WF can achieve target coverage
- Aggregation reduces coverage
- High regime more succeptible to coverage loss
- Spatiotemporal model is not improving coverage
:::

::: {.column width="50%"}
\centering
\small Coverage for day-ahead predictions

\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/batch2025/WF_pred_band_coverage.pdf}
:::

:::


## Coverage under aggregation

::: {.columns}

::: {.column width="50%"}
\centering
\small GB level

\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/batch2025/GB_pred_band_coverage.pdf}
:::

::: {.column width="50%"}
\centering
\small Onshore / offshore

\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/batch2025/GB_pred_band_coverage_tech.pdf}
:::

:::

## Coverage under aggregation by regime
\centering
\small Regime aggregation

\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/batch2025/GB_pred_band_coverage_regime.pdf}



# Out-of-sample locations

## Wind farm prediction bands

::: {.columns}

::: {.column width="50%"}
\centering
\small Linear model (27-01-2025)

\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/batch2025/WF_pred_band_spaceoos_lm_250127_1.pdf}
:::

::: {.column width="50%"}
\centering
\small Spatiotemporal model coarse (27-01-2025)

\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/batch2025/WF_pred_band_spaceoos_st0_m1_250127_1.pdf}
:::

:::

## Wind farm prediction bands

::: {.columns}

::: {.column width="50%"}
\centering
\small 1D SPDE model (27-01-2025)

\includegraphics[width=\linewidth,height=6cm,keepaspectratio]{../fig/batch2025/WF_pred_band_spaceoos_spde1d_250127_1.pdf}
:::

::: {.column width="50%"}
\centering
\small AR1 model (27-01-2025)

\includegraphics[width=\linewidth,height=6cm,keepaspectratio]{../fig/batch2025/WF_pred_band_spaceoos_ar1_250127_1.pdf}
:::

:::

## Aggregated prediction bands

::: {.columns}

::: {.column width="50%"}
\centering
\small GB level (27-01-2025)

\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/batch2025/GB_pred_band_spaceoos_250127.pdf}
:::

::: {.column width="50%"}
\centering
\small Scatter plot oos

\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/batch2025/GB_fit_scatter_oos_space.pdf}
:::

:::


## Unseen locations prediction intervals coverage

::: {.columns}

::: {.column width="50%"}
- Attaining coverage in unseen locations is more challenging
- Isolated locations can lead to high uncertainty
- Location specific events may affect coverage
- Spatiotemporal model is not improving coverage
:::

::: {.column width="50%"}
\centering
\small Coverage for new locations

\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/batch2025/WF_pred_band_coverage_spaceoos.pdf}
:::

:::


## Coverage under aggregation

::: {.columns}

::: {.column width="50%"}
\centering
\small GB level

\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/batch2025/GB_pred_band_coverage_spaceoos.pdf}
:::

::: {.column width="50%"}
\centering
\small Onshore / offshore

\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/batch2025/GB_pred_band_coverage_tech_spaceoos.pdf}
:::

:::

## Coverage under aggregation by regime
\centering
\small Regime aggregation

\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/batch2025/GB_pred_band_coverage_regime_spaceoos.pdf}

# Low wind speed events distribution

## Low wind speed events (insample)

::: {.columns}

::: {.column width="50%"}
\centering
Low wind speed events distribution
\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/batch2025/low_wind_duration_dist_t0_1.pdf}
:::

::: {.column width="50%"}
\centering
Low wind speed events distribution
\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/batch2025/low_wind_duration_qq_t0_1.pdf}
:::

:::

## Low wind speed events (oos time)

::: {.columns}

::: {.column width="50%"}
\centering
Low wind speed events distribution
\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/batch2025/pred_samples_lwe_duration_dist_t0_1.pdf}
:::

::: {.column width="50%"}
\centering
Low wind speed events distribution
\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/batch2025/pred_samples_lwe_duration_qq_t0_1.pdf}
:::

:::

## Low wind speed events (oos space)

::: {.columns}

::: {.column width="50%"}
\centering
Low wind speed events distribution
\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/batch2025/newloc_samples_lwe_duration_dist_t0_1.pdf}
:::

::: {.column width="50%"}
\centering
Low wind speed events distribution
\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/batch2025/newloc_samples_lwe_duration_qq_t0_1.pdf}
:::

:::


# Next steps

## Addressing challenges

- Collapsing uncertainty under aggregation
- Spatiotemporal model is not improving calibration
- Low wind speed events are not well represented in the model samples

# Appendix

## Meshes for spatiotemporal models

::: {.columns}

::: {.column width="50%"}
\centering
Coarse mesh

\includegraphics[width=\linewidth,height=6cm,keepaspectratio]{/home/s2441782/Documents/elexon/extra/mesh_figs/spatial_mesh_coarse_250121.pdf}
:::

::: {.column width="50%"}
\centering
Fine mesh

\includegraphics[width=\linewidth,height=6cm,keepaspectratio]{/home/s2441782/Documents/elexon/extra/mesh_figs/spatial_mesh_fine_250121.pdf}
:::

:::

# Error metrics and performance assessment {-}

## Error metrics

```{r}
  
aggr_summary_tbl <- read.csv("../summaries/calib_metrics_batch2025.csv")
  
aggr_summary_tbl%>%
  mutate(Bias = formatC(Bias, format = "e", digits = 2)) %>% 
  kable(digits = c(3,3,3,3), format = "latex") %>% 
  kable_styling(
    latex_options = c("scale_down")
  )
```

## Error metrics by regime
\vspace{-0.5cm}
```{r regimesummary}
  
aggr_summary_tbl <- read.csv("../summaries/calib_metrics_batch2025_regime.csv")
  
aggr_summary_tbl%>%
  mutate(
    across(matches("Bias"), ~ formatC(., format = "e", digits = 2))
    ) %>%  
  kbl(
  format = "latex",
  col.names = c("Model", rep( c("RMSE", "Bias","MAE"), 3)),
  digits = 3,
  booktabs = TRUE,
  align = c("l", rep("c", 9))
) %>%
  add_header_above(c(
    " " = 1,
    "Low" = 3,
    "Mid" = 3,
    "High" = 3
  )) %>% 
  kable_styling(
    latex_options = c("scale_down")
  )
```

## Error metrics by technology
\vspace{-0.5cm}
```{r techsummary}

aggr_summary_tbl <- read.csv("../summaries/calib_metrics_batch2025_tech.csv")

aggr_summary_tbl%>%
  mutate(
    across(matches("Bias"), ~ formatC(., format = "e", digits = 2))
    # Bias = formatC(Bias, format = "e", digits = 2)
    ) %>% 
  kbl(
  format = "latex",
  col.names = c("Model", rep( c("RMSE", "Bias","MAE"), 2)),
  digits = 3,
  booktabs = TRUE,
  align = c("l", rep("c", 9))) %>%
  add_header_above(c(
    " " = 1,
    "Offshore" = 3,
    "Onshore" = 3
  )) %>% 
  kable_styling(
    latex_options = c("scale_down")
  )
```




# Spatiotemporal calibration model {-}

## Model overview

\begin{equation}
    \mathbb{E}(p^{\mathrm{calib}}_{it})
    = \alpha + \beta_{k} + \beta_m + \beta_h
    + s_p(\hat{p}_{it}) + s_w(w_{it})
    + u_{it},
    \label{eq:bayes_spat}
  \end{equation}

- Calibrated power: corrected output at site *i*, time *t*.  
- $\alpha$: intercept.  
- Fixed effects $\beta_k, \beta_m, \beta_h$: location/season/time structure. 
- ERA5 correction $s_p$: bias adjustment of ERA5 + generic PC estimate.  
- Wind effect $s_w$: nonlinear wind–power correction.  
- Spatiotemporal effect $u_{it}$: Spatial component + AR1 error.


## References

