---
title: "ERA 5 calibration to Elexon power"
author:
  - "S. Gomez$^1$"
institute:
   - $^1$School of Mathematics, University of Edinburgh
institute-short: "UoE"
date: "2026-05-26"
format: 
  beamer:
    theme: Rochester
    colortheme: spruce
    fonttheme: professionalfonts
    fig-cap-location: top
    toc: true
    slide-level: 2
    split-bib: true
    output-file: "wpow_calibration_slides_26-05-26.pdf"
header-includes:
  - \usepackage{graphicx}
fontsize: 12pt
aspectratio: 169
editor: source
bibliography: calib_pow_ref.bib
nocite: "@*"
---

```{r include=FALSE}
require(tidyverse)
require(kableExtra)
require(parallel)
require(dplyr)
require(scales)
require(ggsci)

require(jsonlite)
require(httr)

require(data.table)
require(sf)
library(rnaturalearth)
library(rnaturalearthdata)

require(plotly)
require(ggthemes)
require(arrow)
path <- "~/Documents/elexon"

```

```{r}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo = FALSE, cache = FALSE)
theme_set(theme_bw())
```

# Overview {#sec-intro}


## Updates {#sec-updates .unnumbered .nolink}

### New
\begin{itemize}
  \item Error diagnostics
  \item Spatiotemporal model for calibration
\end{itemize}

### Ongoing

- Model validation
  - Out-of-sample in time
  - Out-of-sample in space
- Consitency under aggregation
- Low wind events

# Model diagnostics

## Diagnostics framework


::: {.columns}

::: {.column width="50%"}

- Performance metrics:
  - RMSE
  - MAE
  - Bias
- Evaluate performance at WF level
- Error under aggregation
  - GB level
  - Onshore / offshore
  - By regime
  - By hour

:::

::: {.column width="50%"}
\centering
\small Sampled days for diagnostics

\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/daily_generation_time_series_sampled_days.pdf}
:::

:::

## Model portfolio

- Wind farm level models:
  - Generic Power Curve (ERA5 + generic PC estimate)
  - Quantile Mapping
  - Linear model 
  - Spatiotemporal model

- Linear model (GB level)

## 

## Error metrics

\small Metrics at wind farm level



## Metrics at GB level

## Technology type

## Regime

## Hour of the day

## Elevation

## Distance to coast

# Linear model {-}

# Spatiotemporal calibration model {-}

## Model overview

\begin{equation}
    \mathbb{E}(p^{\mathrm{calib}}_{it})
    = \alpha + \beta_{k} + \beta_m + \beta_h
    + s_p(\hat{p}_{it}) + s_w(w_{it})
    + u_{it},
    \label{eq:bayes_spat}
  \end{equation}

- Calibrated power: corrected output at site *i*, time *t*.  
- $\alpha$: intercept.  
- Fixed effects $\beta_k, \beta_m, \beta_h$: location/season/time structure. 
- ERA5 correction $s_p$: bias adjustment of ERA5 + generic PC estimate.  
- Wind effect $s_w$: nonlinear wind–power correction.  
- Spatiotemporal effect $u_{it}$: Spatial component + AR1 error.

## Estimated spatial field

\centering

\includegraphics[width=\linewidth,height=7cm,keepaspectratio]{../fig/coarse_spatial_field_median_250810.pdf}

## Standard deviation of the spatial field

\centering

\includegraphics[width=\linewidth,height=7cm,keepaspectratio]{../fig/coarse_spatial_field_sd_250810.pdf}

## Random Effects 
::: {.columns}

::: {.column width="50%"}
\centering
\small Correction by wind speed

\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/wind_effect_coarse_250810.pdf}
:::

::: {.column width="50%"}
\centering
\small Correction by power regime

\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/power_correction_effect_coarse_250810.pdf}
:::

:::

## Hyperparameters

\centering

\includegraphics[width=\linewidth,height=7cm,keepaspectratio]{../fig/hyperparameters_coarse_250810.pdf}

## Comparison scores

```{r}

metrics_table <- read.csv("../summaries/calib_metrics_250810.csv")
# metrics_table <- read.csv("summaries/calib_metrics.csv")

metrics_table %>% 
  mutate(Bias = formatC(Bias, format = "e", digits = 2)) %>% 
  kable(digits = c(3,3,3), format = "latex") %>% 
  kable_styling(
    latex_options = c("scale_down")
  )
```

## Scores by Offshore / Onshore

```{r}
mod_labels <- c(
  "Generic PC",
  "Linear model",
  "Spatio-temporal model",
  "QM"
)
est_cols <- c(
  "norm_power_est0",
  "lm",
  "st",
  "qm"
)

names(mod_labels) <- est_cols
d0_tag <- "250810"
model_df <- st_read(sprintf("../data/calibration_df_%s.gpkg", d0_tag), quiet = TRUE)
df_long <- model_df %>%
  dplyr::select(tech_typ, norm_potential, all_of(est_cols)) %>%
  pivot_longer(
    cols = all_of(est_cols),
    names_to = "model",
    values_to = "estimate"
  )
require(ModelMetrics)
tbl_tech_typ <- df_long %>%
  st_drop_geometry() %>%
  group_by(tech_typ, model) %>%
  summarise(
    RMSE = rmse(actual = norm_potential, predicted = estimate),
    MAE = mae(actual = norm_potential, predicted = estimate),
    Bias = mean(estimate - norm_potential, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(model = mod_labels[model]) %>%
  arrange(tech_typ, desc(RMSE))


tbl_tech_typ %>% 
  mutate(Bias = formatC(Bias, format = "e", digits = 2)) %>% 
  kable(digits = c(3,3,3), format = "latex") %>% 
  kable_styling(
    latex_options = c("scale_down")
  )
```

## Estimates by hour
\centering
\includegraphics[width=\linewidth,height=7cm,keepaspectratio]{../fig/power_estimates_time_series_250810.pdf}

## Errors by hour
\centering
\includegraphics[width=\linewidth,height=7cm,keepaspectratio]{../fig/power_estimates_error_time_series_250810.pdf}


## Model Mesh 

::: {.columns}

::: {.column width="50%"}
\centering
\small Mesh used for spatial calibration model

\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/spatial_mesh_coarse_250810.pdf}
:::

::: {.column width="50%"}
\centering
\small Mesh assessment

\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/spatial_mesh_coarse_assessment_sddev_250810.pdf}
:::

:::

# Time series of data and generic PC estimate
## ERA5 data and generic PC estimate
\centering
\includegraphics[width=\linewidth,height=7cm,keepaspectratio]{../fig/era5_generic_pc_time_series_tech_typ_250810.pdf}

## Observed data

\centering
\includegraphics[width=\linewidth,height=7cm,keepaspectratio]{../fig/observed_power_time_series_tech_typ_250810.pdf}

## Error of generic PC estimate

\centering
\includegraphics[width=\linewidth,height=7cm,keepaspectratio]{../fig/error_time_series_tech_typ_250810.pdf}

## ERA5 wind speed
::: {.columns}

::: {.column width="50%"}
\centering
\small Wind speed at wind farm locations

\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/wind_speed_time_series_250810.pdf}
:::

::: {.column width="50%"}
\centering
\small Observed generation

\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/observed_power_time_series_250810.pdf}
:::

:::

# Same day previous year {-}

## Estimated spatial field

\centering

\includegraphics[width=\linewidth,height=7cm,keepaspectratio]{../fig/coarse_spatial_field_median_240810.pdf}

## Standard deviation of the spatial field

\centering

\includegraphics[width=\linewidth,height=7cm,keepaspectratio]{../fig/coarse_spatial_field_sd_240810.pdf}

## Random Effects 
::: {.columns}

::: {.column width="50%"}
\centering
\small Correction by wind speed

\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/wind_effect_coarse_240810.pdf}
:::

::: {.column width="50%"}
\centering
\small Correction by power regime

\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/power_correction_effect_coarse_240810.pdf}
:::

:::

## Hyperparameters

\centering

\includegraphics[width=\linewidth,height=7cm,keepaspectratio]{../fig/hyperparameters_coarse_240810.pdf}

## Comparison scores

```{r}

metrics_table <- read.csv("../summaries/calib_metrics_240810.csv")
# metrics_table <- read.csv("summaries/calib_metrics.csv")

metrics_table %>% 
  mutate(Bias = formatC(Bias, format = "e", digits = 2)) %>% 
  kable(digits = c(3,3,3), format = "latex") %>% 
  kable_styling(
    latex_options = c("scale_down")
  )
```

## Scores by Offshore / Onshore

```{r}
mod_labels <- c(
  "Generic PC",
  "Linear model",
  "Spatio-temporal model",
  "QM"
)
est_cols <- c(
  "norm_power_est0",
  "lm",
  "st",
  "qm"
)

names(mod_labels) <- est_cols
d0_tag <- "240810"
model_df <- st_read(sprintf("../data/calibration_df_%s.gpkg", d0_tag), quiet = TRUE)
df_long <- model_df %>%
  dplyr::select(tech_typ, norm_potential, all_of(est_cols)) %>%
  pivot_longer(
    cols = all_of(est_cols),
    names_to = "model",
    values_to = "estimate"
  )
require(ModelMetrics)
tbl_tech_typ <- df_long %>%
  st_drop_geometry() %>%
  group_by(tech_typ, model) %>%
  summarise(
    RMSE = rmse(actual = norm_potential, predicted = estimate),
    MAE = mae(actual = norm_potential, predicted = estimate),
    Bias = mean(estimate - norm_potential, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(model = mod_labels[model]) %>%
  arrange(tech_typ, desc(RMSE))


tbl_tech_typ %>% 
  mutate(Bias = formatC(Bias, format = "e", digits = 2)) %>% 
  kable(digits = c(3,3,3), format = "latex") %>% 
  kable_styling(
    latex_options = c("scale_down")
  )
```

## Estimates by hour
\centering
\includegraphics[width=\linewidth,height=7cm,keepaspectratio]{../fig/power_estimates_time_series_240810.pdf}

## Errors by hour
\centering
\includegraphics[width=\linewidth,height=7cm,keepaspectratio]{../fig/power_estimates_error_time_series_240810.pdf}


## Model Mesh 

::: {.columns}

::: {.column width="50%"}
\centering
\small Mesh used for spatial calibration model

\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/spatial_mesh_coarse_240810.pdf}
:::

::: {.column width="50%"}
\centering
\small Mesh assessment

\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/spatial_mesh_coarse_assessment_sddev_240810.pdf}
:::

:::

# Time series of data and generic PC estimate {-}
## ERA5 data and generic PC estimate
\centering
\includegraphics[width=\linewidth,height=7cm,keepaspectratio]{../fig/era5_generic_pc_time_series_tech_typ_240810.pdf}

## Observed data

\centering
\includegraphics[width=\linewidth,height=7cm,keepaspectratio]{../fig/observed_power_time_series_tech_typ_240810.pdf}

## Error of generic PC estimate

\centering
\includegraphics[width=\linewidth,height=7cm,keepaspectratio]{../fig/error_time_series_tech_typ_240810.pdf}

## ERA5 wind speed
::: {.columns}

::: {.column width="50%"}
\centering
\small Wind speed at wind farm locations

\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/wind_speed_time_series_240810.pdf}
:::

::: {.column width="50%"}
\centering
\small Observed generation

\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/observed_power_time_series_240810.pdf}
:::

:::


# Appendix 

# Project Goal {-}

- Correct ERA5 wind power biases using observed generation data.
- Reduce error metrics (RMSE/MAE/bias).
- Ensure spatially consistent estimates that generalise out-of-sample.
- Capture low-wind events: frequency, duration, and spatial extent.

## Spatial correlation {-}
::: {.columns}

::: {.column width="50%"}
\small Correlation for Elexon generation

\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/norm_pot_corr_dist.pdf}
:::

::: {.column width="50%"}
\small Correlation for error after calibration

\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/error0_corr_dist.pdf}
:::

:::

# Calibration of generic power curve estimate {-}

## GB level aggregation
::: {.columns}

::: {.column width="50%"}
\footnotesize
- Nearest ERA5 grid for wind speed
- Vertical interpolation using a power law
- Generic Power Curve (onshore, offshore)
- Normalised ower aggregated at GB level
  \begin{align*}
  p_{\text{tot},t} &= \sum_i p_{it} / \sum_i C_i 
  \end{align*}
- Hourly resolution
- Potential generation (ex curtailment)
:::

::: {.column width="50%"}
\centering
\small Scatter aggregated power
\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/gb_est0_hexbin_all.pdf}
:::

:::

## Linear model

::: {.columns}

::: {.column width="50%"}
\footnotesize
- Using step AIC criteria selection: 
  - $k$: Type (offshore / onshore)
  - $m$: month, $h$: hour
  - $w_t$: wind speed


\begin{equation*}
p^{calib}_t = \alpha_{k,m,h} + \beta_{k,m} \hat p_{t} + \sum_{j=1}^3 \phi_j w_t^3
\end{equation*}
:::

::: {.column width="50%"}
\centering
\small Calibration comparison
\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/gb_lmcalib_hexbin_all.pdf}
:::

:::

## Autoregressive error

::: {.columns}

::: {.column width="50%"}
\footnotesize
- Using step AIC criteria selection: 
  - $k$: Type (offshore / onshore)
  - $m$: month, $h$: hour
  - $w_t$: wind speed
  - $u_t$: AR1

\begin{equation*}
p^{calib}_t = \alpha_0 + \sum \alpha_e + \beta_{k} \hat p_{t} + f_w(w_t) + u_t
\end{equation*}

where $\alpha_0$ is the intercept, the $\alpha_e$ $e \in {k,m,h}$ represent random effects (cyclic for month and hour), $\beta_k$ is a random slope, $f_w$ is smooth effect (RW2). 
:::

::: {.column width="50%"}
\centering
\small Calibration comparison
\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/gb_ar1calib_hexbin_all.pdf}
:::

:::

## Hierarchical Bayesian model

::: {.columns}

::: {.column width="50%"}
\footnotesize
- Turning muliplicative corrections into smooth functions


\begin{equation*}
p^{calib}_t = \alpha + \beta_k + \beta_m + \beta_h + s_p(p_{t}) + s_w (w_t)
\end{equation*}

where $\beta_k$, $\beta_m$, $\beta_h$ are random effects and $s_p$ and $s_w$ are smooth functions
:::

::: {.column width="50%"}
\centering
\small Calibration comparison
\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/gb_lmbrucalib_hexbin_all.pdf}
:::

:::

## Quantile Mapping

::: {.columns}

::: {.column width="50%"}
\footnotesize
- Quantile mapping aims to replicate observed distribution

\begin{equation*}
p^{calib}_t = F_\text{obs} (F_\text{mod}^{-1}(p_{t}))
\end{equation*}

where $F_\text{mod}$ and $F_\text{obs}$ are the CDF of the power estimate and observed values
:::

::: {.column width="50%"}
\centering
\small Calibration comparison
\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/gb_qmcalib_hexbin_all.pdf}
:::

:::

## Comparison scores

```{r}

metrics_table <- read.csv("../summaries/calib_metrics.csv")
# metrics_table <- read.csv("summaries/calib_metrics.csv")

metrics_table %>% 
  mutate(Bias = formatC(Bias, format = "e", digits = 2)) %>% 
  kable(digits = c(3,3,3), format = "latex") %>% 
  kable_styling(
    latex_options = c("scale_down")
  )
```

# Power curve modelling {-}

## Power curve modelling
::: {.columns}

::: {.column width="50%" }
### Components

- Power curve estimate from data
- Probability of zero generation
- Penalisation to make power curve resemble manufaturer's PC

:::

::: {.column width="50%"}
Using a three-likelihood approach for power curve estimation:

- **1st likelihood:** Observed power is modelled as a smooth function of wind speed
- **2nd likelihood:** Zero inflated component
- **3rd likelihood** Penalisation towards a generic PC

:::

:::

## Power model overview

\footnotesize

1. $\pi_{it}=\pi(w, i) = \operatorname{Pr}(P_{it} = 0)$ depends of wind speed and location
\begin{equation*}
L^{(0)}(\theta^0| \mathbf{p}, \mathbf{w}, \mathbf{i}) = \prod_{t} \left[ \pi_i(w_{it}, i)^{\mathbb{1}(P_{it} = 0)} \times (1 - \pi_i(w_{it}, i))^{\mathbb{1}(P_{it} > 0)} \right]
\end{equation*}
2. Generation when $P_{it} > 0$ also depends on wind speed / location. Learned as a smooth function from data: $s^{\mathrm{pc}}(w, i) \in (0,1)$
\begin{align*}
s^{\mathrm{pc}}(w_{it},i) &= \text{E}(P_{it}| P_{it}>0, w_{it}, i) = \mu^+_{it}\\
L^{(\beta)}(\theta^{\mathrm{pc}}| \mathbf{p}, \mathbf{w}, \mathbf{i}) &= \prod_{t: P_{it} > 0} \text{Beta}\left(P_{it} | \mu^+_{it}, \phi \right),
\end{align*}
where $\theta^{0}$ and $\theta^{\mathrm{pc}}$ represents the parameters of the smooth functions, $\pi(w,i)$ and $s^{\mathrm{pc}}$, and $\phi$ the precision of the Beta distribution.

<!-- ## Power model overview
\begin{align*}
\mu_{it} &= E[P_{it}|w_{it}, i] \\
&=
(1 - \pi_i(w_{it}, i)) \times s_{\mathrm{pc}}(w_{it}, i)\\
s_{\mathrm{pc}}(w_{it}, i) &\approx \hat{PC}_i(w_{it})
\end{align*} -->

## Penalisation component

3. Finally, we penalise $s^{\mathrm{pc}}$ to be close to our generic power curve estimate, $\hat{PC}_i(w)$: 
$d_{it}=|\hat{PC}_i(w_{it}) - s^{\mathrm{pc}}(w_{it},i)|$ should be small. 

- Or maximise the negative of $\tau_{\mathrm{ps}} \sum d^2_{it}$ for a certain penalisation parameter $\tau_{\mathrm{ps}}$.
- Adding that penalty term is similar to assume $d_{it} \sim N(0, 1/\tau_{\mathrm{ps}}^{-1})$
\begin{equation*}
\max_{\theta} \quad l^{(0)}(\theta) + l^{(\beta)}(\theta) - \tau_{\mathrm{ps}} \sum_{i,t} d^2_{it}
\end{equation*}
where $\theta = (\theta^0,\theta^{\mathrm{pc}}, \tau_{\mathrm{ps}})$ represents the parameters of both likelihoods, plus the penalisation parameter. 

## Power curve model equations
\tiny
Model is a Zero Inflated Beta (ZIB) with penalisation component:
\begin{align*}
%%
&\textbf{Zero--probability model (Bernoulli)} \nonumber \\[4pt]
\eta^{(0)}_{it}
  &= \alpha_{\mathrm{zero}}
     + f_{\mathrm{zero}}\!\left(w_{it},\,i\right), \\[4pt]
Z_{it} \mid \eta^{(0)}_{it} 
  &\sim \mathrm{Bernoulli}(\pi_{it}),
  \qquad 
  \pi_{it} = \frac{1}{1 + \exp\left\{-\eta^{(0)}_{it}\right\}},
\\[10pt]
%%
&\textbf{Positive--output model (Beta)} \nonumber \\[4pt]
\eta^{(\beta)}_{it}
  &= \alpha_{\mathrm{pc}}
     + f_{\mathrm{pc}}\!\left(w_{it},\,i\right), \\[4pt]
P_{it}^{\text{obs}} \mid \eta^{(\beta)}_{it}
  &\sim \mathrm{Beta}(\mu^+_{it},\phi),
  \qquad
  \mu^+_{it} = \frac{1}{1 + \exp\left\{-\eta^{(\beta)}_{it}\right\}},
\\[10pt]
%%
&\textbf{Pseudo--likelihood model (Gaussian)} \nonumber \\[4pt]
\hat{PC}_i(w_{it}) \mid \eta^{(\beta)}_{it}
  &\sim \mathcal{N}\!\left(
      \eta^{(\beta)}_{it},
      \tau_{\mathrm{ps}}^{-1}
    \right),
  \qquad
    \tau_{\mathrm{ps}} = \text{precision or penalisation}.
\end{align*}

## Details on smooth components
\footnotesize
- The smooth components $f_{\mathrm{zero}}(w,i)$ $f_{\mathrm{pc}}(w,i)$ is modelled as a 1D SPDE over wind speed with a uniform correlation matrix across locations $i = 1, \ldots, I$.
\begin{equation*}
f(w, i) = \sum_{i=1}^{I} \beta_{i} \psi_k(w),
\end{equation*}
where $\psi_k(w)$ are basis functions defined over wind speed and $\beta_i$ are coefficients
The correlation structure of $f(w, i)$ assumes a Matern covariance function:
Let $w_{s}$, $w_h$ represent two wind speed values in location $i$, and $d = |w_s - w_t|$ their distance, then:
\begin{align*}
\operatorname{Cor}(d) &= \operatorname{Cor}(f(w_s,i), f(w_t, i))\\
& = \frac{1}{2^{\nu-1} \Gamma(\nu)}  (\kappa d)^\nu K_\nu (\kappa d), \quad \alpha = \nu + d/2,
\end{align*}

## Details on smooth components
- Additionally a uniform correlation matrix is assumed across locations $j,k \in \{ 1, \ldots, I\}$.
That is: 
\begin{equation*}
\operatorname{Cor}(f(w,j), f(w, k))  = \rho, \quad j \neq k
\end{equation*}
which helps to pool information across locations, benefiting locations with less data.

## Power estimate

This leads to the following power estimate at time $t$ and location $i$:
\begin{equation*}
\hat P_{it} = \begin{cases}
0, & \text{with probability } \pi_{it},\\
\mu^+_{it}, & \text{with probability } 1 - \pi_{it}.
\end{cases}
\end{equation*}

Which leads to the expected power estimate:
\begin{equation*}
E[P_{it}|w_{it}, i] = (1 - \pi_{it}) \times \mu^+_{it}
\end{equation*}

## Power curve model estimates

::: {.columns}

::: {.column width="50%" }

- Starting models with 
  - 1Y of data (2024)
  - 25 wind farms
  - Hourly data
  - Excluding outages listed in REMIT

:::

::: {.column width="50%"}
\centering
\small Expected Power curve

\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/ZIBspdePen0.10free_24_wfsamp_EPC.pdf}
:::

:::
## Power curve model estimates

::: {.columns}

::: {.column width="50%" }
\centering
\small Power curve estimate

\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/ZIBspdePen0.10free_24_wfsamp_curve.pdf}

:::

::: {.column width="50%"}
\centering
\small Probability of zero generation

\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/ZIBspdePen0.10free_24_wfsamp_zeroProb.pdf}
:::

:::

## Alternative versions

::: {.columns}

::: {.column width="50%" }
\footnotesize

- Beta model with RW2 power curve (B-RW2)
- Beta model with 1D SPDE power curve (B-SPDE)
- ZIB model with 1D SPDE power curve (ZIB-SPDE)
- ZIB model with 1D SPDE power curve penalised (ZIB-SPDE-P)

Zero Inflated models better capture low wind power regions.


:::

::: {.column width="50%"}
\centering
\small Power curve estimates under different models

\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/comparison_24_1wf_curve.pdf}
:::

:::

## ZIB model vs observed data
::: {.columns}

::: {.column width="50%" }
\footnotesize

- Power curve modelling highly improves the calibration
- ZIB models deal better with the tail regions
- Penalised version actually pulls estimate toward Generic PC
- We could pull estimate instead to a 5 parameter curve estimate

:::

::: {.column width="50%"}
\centering
\small Power curve models vs observed data

\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/gbwfsamp_pcmodel_hexbin_2024.pdf}
:::

:::
## Error metrics
Error metrics calculated on 30% of the sampled wind farms data.
```{r}
require(ModelMetrics)
mod_names <- c("B-RW2", "B-SPDE", rep("ZIB-SPDE", 3), rep("ZIB-SPDE-P", 3))
mod_names_short <- mod_names %>% unique()
aggr_pcmod_df <- read_parquet("../data/pcmodel_wfsamp_aggr_test.parquet")
# aggr_pcmod_df <- read_parquet("data/pcmodel_wfsamp_aggr_test.parquet")
mod_labels <- c("Generic PC", mod_names_short)
est_cols <- c(
  "norm_power_est0",
  "pred_rw2",
  "pred_spde",
  "pred_zib",
  "pred_zibp"
)
names(mod_labels) <- est_cols
df_long <- aggr_pcmod_df %>%
  select(norm_potential, all_of(est_cols)) %>%
  pivot_longer(
    cols = all_of(est_cols),
    names_to = "model",
    values_to = "estimate"
  )
metrics_table <- df_long %>%
  group_by(model) %>%
  summarise(
    RMSE = rmse(actual = norm_potential, predicted = estimate),
    MAE = mae(actual = norm_potential, predicted = estimate),
    Bias = mean(estimate - norm_potential, na.rm = TRUE),
    .groups = "drop"
  ) %>% 
  mutate(model = mod_labels[model])
metrics_table %>% 
  kable(digits = 4, format = "latex") %>% 
  kable_styling(
    latex_options = c("scale_down")
  )
```

## Pooling data for cut-off estimation
::: {.columns}
::: {.column width="50%" }
\small
- Not enough data to estimate at wind farm level
- Even using all 5 years of data (2019-2024), only a few farms show cut-off behaviour

\centering
Pooling data at high wind speeds
\footnotesize
```{r}

wf_pattern_aggr <- read.csv("../data/cutoff_wind_farms_pattern.csv")

wf_pattern_aggr %>% 
  select(-cutoff_pattern) %>% 
  kable(
    col.names = c("wind speed", "prop. P=0", "prop. P<=0.02", "n obs."),
    digits = c(0,2,2,0),
    format = "latex",
    big.mark = ","
  ) %>% 
  kable_styling(
    latex_options = c("scale_down")
  )
```

:::

::: {.column width="50%"}
\small
\centering
Tail behaviour of power curve
\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/cutoff_wind_farms.pdf}
:::
:::

## Model to estimate cut-off by wind speed

::: {.columns}
::: {.column width="50%" }
\centering
Hyperparameters of the cut-off model
\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/cutoffspde_cut_wfsamp_hyper.pdf}
:::

::: {.column width="50%"}
\small
\centering
Probability of zero generation at high wind speeds
\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/cutoffspde_cut_wfsamp_zeroProb.pdf}
:::
:::


# Power Curve fitting{-}

## Classic parametric shape

::: {.columns}

::: {.column width="50%"}

\footnotesize
The five-parameter formula (Lydia et.al, 2013):

\begin{equation*}
  P(u) = D+ \frac{A-D}{(1+(u/C)^B)^G} 
\end{equation*}
where: $A$ and $D$ are the upper and lower asymptotes, $C$ controls the inflection point, $B$ is related to the slope at inflection point, and $G$ controls the asymmetry.

:::

::: {.column width="50%"}
\centering
\small 5 parameters power curve
```{r fig.width = 4.5, fig.height = 3.5}
pc5param <- function(x, A, B, C, D, G) {
  D + (A - D) / (1 + (x / C)^B)^G
}
D <- 0
A <- 1
B = -15
G = 0.4
C = 7
w <- seq(0, 25, length.out = 100)
p <- pc5param(w, A, B, C, D, G)

data.frame(
  wind = w,
  power = p
) %>%
  ggplot(aes(wind, power)) +
  geom_line(col = blues9[7])
```

:::

:::

## Parametric shape for cut-off

::: {.columns}

::: {.column width="50%"}

\footnotesize
Adding extra parameters we can model de decay after cut-off:

\begin{equation*}
  P(u) = D+ \frac{A-D}{(1+(u/C_1)^{B_1})^G(1+(u/C_2)^{B_2})^G} 
\end{equation*}
where: $A$ and $D$ are the upper and lower asymptotes, $C_1$ is the first inflection point, $C_2$ is related to the cut-off, $B_1$ is related to the slope at inflection point, $B_2$ to the slope during cut-off, and $G$ controls the asymmetry.

:::

::: {.column width="50%"}

\centering
\small 7 parameters power curve

```{r fig.width = 4.5, fig.height = 3.5}
pc7param_r <- function(x, A_raw, B1_raw, C1, D, G_raw, C2, B2_raw) {
  B1 <- -exp(B1_raw) # always negative
  B2 <- exp(B2_raw) # always positive
  A <- D + exp(A_raw) # ensures A > D
  G <- exp(G_raw) # ensures G > 0

  D +
    (A - D) /
      (1 + (x / C1)^B1)^G /
      (1 + (x / C2)^B2)^G
}
D <- 0
A <- 1
B1 = log(15)
B2 = log(100)
G = log(0.4)
C1 = 7
C2 = 22
w <- seq(0, 30, length.out = 100)
p <- pc7param_r(w, A, B1, C1, D, G, C2, B2)

data.frame(
  wind = w,
  power = p
) %>%
  ggplot(aes(wind, power)) +
  geom_line(col = blues9[7])
```

:::

:::


## Parametric fit for Scottish wind farms

::: {.columns}

::: {.column width="50%"}

\footnotesize

- We can fit the parametric shape using optim
- As long as there is a reasonable amount of points showing the cut-off optim can converge

\tiny
\centering
Estimates from data

```{r}
pre_path <- "/home/s2441782/ownCloud-s2441782@datasync.ed.ac.uk/projects/proj2/prob-scen-spatial"
solution1 <- read.csv(file.path(pre_path,"data/pc_7pars_wfsamp_24b.csv")) %>%
  rename(par = X)
solution1 %>%
  kable(digits = 2, format = "latex") %>%
  kable_styling(full_width = FALSE)
```
:::

::: {.column width="50%"}
\centering
\small Data and power curve estimate

\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/scottish_wfsamp_24_pc-est.png}
:::

:::





## Next Steps


- Spatiotemporal calibration model
- Non-stationary covariance model
- Comparison against benchmarks


<!-- - Add CRPS and coverage metrics -->

## Data sources 
::: {.columns}

::: {.column width="50%"}
### Wind speed
- ERA5 at wind farms
  - Hourly data
  - Spatial resolution 0.25° x 0.25° 
  - 10m and 100m heights

:::

::: {.column width="50%"}
### Wind power
- Elexon BMU data (since 2019)
  - Half hourly data
  - Generation, curtailment, potential, capacity
  - Outage data (REMIT)
- REPD database
  - Location, turbine height, capacity
:::

:::


# Previous research {-}

## Power curve modelling

::: {.columns}

::: {.column width="50%"}

### Binning method
\begin{equation*}
  P_i = \frac{1}{n_i} \sum_{j=1}^{n_i} P_{ij}
\end{equation*}
where: $P_{ij}$ is the $j$ th power observation in bin $i$ and
$n_i$ no. of observations in bin $i$

:::

::: {.column width="50%"}
### Logistic
\begin{equation*}
P(u) = a \frac{1+ m \exp(-u/\tau)}{1+ n \exp(-u/\tau)}
\end{equation*}
where $a$ represents the upper asymptote, $n, m$ shape the lower asymptote, and $\tau$ controls the transition. 
:::

:::

## Power curve modelling

::: {.columns}

::: {.column width="50%"}

### 5 parameter curve
\begin{equation*}
  P(u) = D+ \frac{A-D}{(1+(u/C)^B)^G} 
\end{equation*}
where: $A$ and $D$ are the upper and lower asymptotes, $C$ is the inflection point, $B$ the slope at inflection point, adn $G$ controls the asymmetry.

:::

::: {.column width="50%"}

:::

:::

## Reanalysis data to quantify extreme wind power statistics
::: {.columns}

::: {.column width="50%"}
\footnotesize

D. Cannon, D. Brayshaw, et. al (2015)

- MERRA wind speed validated with MIDAS
- Vertical interpolation with a logarithmic change
- Calibrated power curves based on manufacturers PC
- Use that to analys extreme low and high levels, and ramps
:::

::: {.column width="50%"}
\centering
\small $\Delta$CF comparison 
\includegraphics[width=\linewidth]{../fig_extra/2013_cannon_delta_comparison.png}
:::

:::

## Balancing energy
::: {.columns}

::: {.column width="50%"}
\footnotesize
H. Thornton, D. Brayshaw (2020) analyse the relationship between weather, energy demand, and wind power.

- ERA Interim wind speed
- Cubic power curve with air density correction
- CF compared with GB average from other studies
- Seasonal effects on Demand and 

:::

::: {.column width="50%"}
\centering
\small $\Delta$CF comparison 
\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig_extra/2020_thornton_CF_vs_demand.png}
:::

:::

## Analysis of extreme wind droughts
::: {.columns}

::: {.column width="50%"}
\footnotesize
Panit Potisomporn, C. Vogel (2024) perform a extreme value analysis of wind droughts in GB.

- Use ERA5 wind speeds calibrated to MIDAS with QM
- Build a ML algorithm that learns how to extrapolate wind speed from 10m to hub height
- Use a 5 parameter logistic function to model power curve
- Model energy losses with factors by type.

:::

::: {.column width="50%"}
\centering
\small CF calibration 
\includegraphics[width=\linewidth]{../fig_extra/2024_potisomporn_CF_comparison.png}
:::

:::

# ERA5 + generic power curve error analysis {-}

## Error by Onshore / Offshore
::: {.columns}

::: {.column width="50%"}
\centering
\small Errors by observed power

\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../../calibration_power_main_doc/fig_errors/gb_tech+_genPC_resid_2Ddensity.pdf}
:::

::: {.column width="50%"}
\centering
\small Error density

\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../../calibration_power_main_doc/fig_errors/gb_tech_genPC_resid_density.pdf}
:::

:::

## Errors through time
::: {.columns}

::: {.column width="50%"}
\centering
\small Error by hour

\includegraphics[width=\linewidth,height=7cm,keepaspectratio]{../../calibration_power_main_doc/fig_errors/gb_tech+hour_genPC_resid_boxplot.pdf}
:::

::: {.column width="50%"}
\centering
\small Error by month

\includegraphics[width=\linewidth,height=7cm,keepaspectratio]{../../calibration_power_main_doc/fig_errors/gb_tech+month_genPC_resid_boxplot.pdf}
:::

:::



# Challenges in spatial wind calibration {-}

- Non-stationarity 
  - (space) spatially varying coefficients
  - (wind direction) anisotropy
  - (time) regimes / storms and other weather events

- Physically informed model 
  - Regularisation / penalisation

- Data dimensionality reduction
  - Aggregate time dimension (day)


# Non-stationarity approaches found in the literature {-}


## Regression with PDE regularisation
::: {.columns}

::: {.column width="50%"}
\footnotesize
- Tomasetto et al.(2024) model anisotropy through a physics informed regression model
- Includding physical knowledge through a PDE regularisation term
\centering
\includegraphics[width=\linewidth,height=3cm,keepaspectratio]{../fig_extra/2024_tomasetto_s-pde.png}
:::

::: {.column width="50%"}
\centering
\small 
Sea Surface Temperature (SST)
\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig_extra/2024_tomasetto_temp.png}
:::

:::

## Explanatory variables in the covariance
::: {.columns}

::: {.column width="50%"}
\footnotesize
- Ingebrigtsen, Lindgren, and Steinsland (2014)
- Regression on ranke and precision parameters of spatial term

\begin{align}
  \log \tau(\mathbf{s}) &= b_0^{(\tau)}(\mathbf{s}) +
    \sum_{k=1}^p b_k^{(\tau)}(\mathbf{s})\,\theta_k,
  \label{eq:logtau}\\
  \log \kappa(\mathbf{s}) &= b_0^{(\kappa)}(\mathbf{s}) +
    \sum_{k=1}^p b_k^{(\kappa)}(\mathbf{s})\,\theta_k.
  \label{eq:logkappa}
\end{align}

:::

::: {.column width="50%"}
\centering
\small 
Field and correlation at two points
\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig_extra/2018_krasinksi_cov_regress.png}
:::

:::



## Variables relevant to non-stationarity in wind
::: {.columns}

::: {.column width="50%"}
\footnotesize
- Wind dynamics are non-stationary in space and time
- Weather regimes, storms, other events
- Circulation patterns (NAO, AO, EA, SCAN)
- NAO + leads to higher wind speeds UK and more variability

:::

::: {.column width="50%"}
\centering
\small 
North Atlantic Oscillation (NAO)
\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig_extra/2026_noaa_nao.png}
:::

:::

## Cirtulation patterns and wind power

::: {.columns}

::: {.column width="50%"}
\centering
\footnotesize
North Atlantic Oscillation Index (NAO)
\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/nao_generation.pdf}

:::

::: {.column width="50%"}
\centering
\footnotesize
Artic Oscillation Index (AO)
\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/ao_generation.pdf}
:::

:::

## Cirtulation patterns and wind power

::: {.columns}

::: {.column width="50%"}
\centering
\footnotesize
East Atlantic Index (EA)
\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/ea_generation.pdf}
:::

::: {.column width="50%"}
\centering
\footnotesize
Scandinavian Oscillation Index (SCAN)
\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/scan_generation.pdf}
:::

:::



## Cirtulation patterns and error in wind power estimates

::: {.columns}

::: {.column width="50%"}
\centering
\footnotesize
North Atlantic Oscillation Index (NAO)
\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/nao_error.pdf}

:::

::: {.column width="50%"}
\centering
\footnotesize
Artic Oscillation Index (AO)
\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/ao_error.pdf}
:::

:::

## Cirtulation patterns and error in wind power estimates

::: {.columns}

::: {.column width="50%"}
\centering
\footnotesize
East Atlantic Index (EA)
\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/ea_error.pdf}
:::

::: {.column width="50%"}
\centering
\footnotesize
Scandinavian Oscillation Index (SCAN)
\includegraphics[width=\linewidth,height=5cm,keepaspectratio]{../fig/scan_error.pdf}
:::

:::

## References

